Revisiting the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation using complex Empirical Orthogonal Functions

### 2022-03-23 analysis/paper/shceof.Rmd
Revisión

The Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope [@irving2015].
The latter makes possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
With this approach, @irving2015 showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.

@@

The Oceanic Niño Index [ONI, @bamston1997] comes from NOAA's Climate Prediction Center and and the Dipole Mode Index [DMI, @saji2003] from Global Climate Observing System Working Group on Surface Pressure.
In both cases, the monthly time series averaged seasonally to produce seasonal time series.

@@

The study is restricted to the spring season, defined as the September-October-November (SON) trimester.
We compute seasonal means for the different variables, averaging monthly values weighted by the number of days in each month.
We use 200 hPa level to represent the high troposphere and 50 hPa to represent the lower stratosphere.

@@

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according [@baldwin2009].
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2022.
Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated -- in agreement with @mo2001 -- as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height.

@@

In standard way of applying EOF analysis zonal waves may appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase [@horel1984].
Figure \@ref(fig:eof-naive) shows the four eading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
A similar EOF structure can be seen in 200 hPa (not shown).
Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

ON the other hand, complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, associated with phase-varying structures [@horel1984].
This method involves computing EOF on the analytic representation of the original field.
That representation is a complex field in which the real part is the original data and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e. its Hilbert transform.
The Hilbert transform is usually understood in terms of time-varying signal.
However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.

@@

The phase of principal components is defined up to an additive constant.
For real principal components, this constant can be either 0 or $\pi$, corresponding to a change in sign.
For complex principal components, it can be any number between 0 and $2\pi$ [@horel1984].
Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
This procedure does not create a spurious correlation, it only takes a known relationship and aligns it with a specific phase.
For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
This also nearly minimises the correlation with the imaginary part.
For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.

@@

Lastly, the cEOF1 doesn't have a significant signal associated with 2-metre temperature or precipitacion (Fig. \@ref(fig:pp-temp-cap-1)).

@@

In this study we assessed the Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation.
For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions.
This allowed us to characterise both the amplitude and phase of planetary waves that are not perfectly organised as sinusoidal waves with constant amplitude at each latitude circle.

>>>>

The Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope [@irving2015].
The latter makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
With this approach, @irving2015 showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.

@@

The Oceanic Niño Index [ONI, @bamston1997] comes from NOAA's Climate Prediction Center and the Dipole Mode Index [DMI, @saji2003] from Global Climate Observing System Working Group on Surface Pressure.

@@

The study is restricted to the spring season, defined as the September-October-November (SON) trimester.
We compute seasonal means for the different variables, averaging monthly values weighted by the number of days in each month.
We use the 200 hPa level to represent the high troposphere and 50 hPa to represent the lower stratosphere.

@@

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009].
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by @campitelli2022.
Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated, in agreement with @mo2001, as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.

@@

In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerated) EOFs representing similar patterns but shifted in phase [@horel1984].
Figure \@ref(fig:eof-naive) shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with shorter wavenumber and four centres of action shifted by 1/4 wavelength.
A similar EOF structure can be seen in 200 hPa (not shown).
Since each pair of EOFs seems to represent the same phase-varying structure, it would be desirable to combine them into a single pattern described by amplitude and phase.

Complex Empirical Orthogonal Functions (cEOF) are a useful method for characterising zonal waves, associated with phase-varying structures [@horel1984].
This method involves computing EOF on the analytic representation of the original field.
That representation is a complex field in which the real part is the original data and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e. its Hilbert transform.
The Hilbert transform is usually understood in terms of time-varying signal.
However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.

@@

The phase of cEOFs is defined up to an additive constant.
For real EOFs, this constant can be either 0 or $\pi$, corresponding to a change in sign.
For cEOFs, it can be any real number between 0 and $2\pi$ [@horel1984], corresponding to rotations in the complex plane.
Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
This procedure does not create a spurious correlation, it only takes an existing relationship and aligns it with a specific phase.
For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
This also nearly minimises the correlation with the imaginary part.
For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.

@@

@@

In this study we assessed the Austral Spring Extratropical Southern Hemisphere zonally asymmetric circulation.
For this purpose, two complex indices were derived using Complex Empirical Orthogonal Functions.
This allowed us to characterise both the amplitude and phase of planetary waves that are not perfectly organised as sinusoidal waves with constant amplitude at each latitude circle.
